experDesign: helping performing experiments on collections

### 2021-04-09 paper.Rmd
Update paper with some` information about the other packages

# Statement of need

@@

Between the designing of an experiment and the collection of the samples many things can happen. If something goes wrong and some samples are missing they can increase confounding variables or even make the study useless. However, few tools exists to prevent generating a batch effect.

To prevent the batch effect there can be two options: randomization and/or replication. Whether the practitioner goes with randomization or replication they must take into account the blocking design from the beginning or they might. For instance, if cases and controls are run in different batches, sample variation can be entirely confounded by this [@chen2011]. The samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch.

The package experDesign for R is an extension to R that all[@rcoreteam2014]

Several functions of experDesign allow this. The main function `design`distribute the samples on multiple batches so that each variable is equally homogeneous in each batch. If the experiment is carried out on a specific spatial distribution the `spatial` function takes into consideration the position of the samples to distribute them homogeneously also by position.

>>>>

# Introduction

@@

Between the designing of an experiment and the collection of the samples many things can happen. If something goes wrong and some samples are missing they can increase confounding variables or even make the study useless.

To prevent the batch effect there can be two options: randomization and/or replication. Whether the practitioner goes with randomization or replication they must take into account the blocking design from the beginning or they might. For instance, if cases and controls are run in different batches, sample variation can be entirely confounded by this [@chen2011]. The samples can be properly randomized to minimize the batch effect by looking how the variables distribute on each batch, on what it is know as randomized blocked/ stratified sampling experimental design.

# State of the art

There are some tools to prevent generating a batch effect on the R language from multiple fields and areas, but specially on the biological research [@rcoreteam2014]. They have some caveats that limit their application to some cases and to our knowledge there hasn't been a comparison of these different methods.

*OSAT*, on [Bioconductor](https://bioconductor.org/packages/OSAT/), first allocates the samples on each batch according to a variable and later shuffles the samples on each batch to randomize the other variables [@yan2012]. This algorithm relies on categorical variables and cannot use numeric ones such as age or time-related variables unless they are treated as categorical.

*minDiff*, on [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, on [CRAN](https://cran.r-project.org/package=anticlust), [@usingan] divided elements into similar parts, ensuring similarity between groups by enforcing heterogeneity within group. Conceptually it is similar to reversing the clustering methods k-means and cluster editing, and as them it must use numeric variables.

Recently, *Omixer*, a new package on [Bioconductor](https://bioconductor.org/packages/Omixer/), has been made available [@sinke2021]. It tests if the random assignments are homogeneous transforming all variables to numeric and using the Kendall's correlation if there are more than 5 samples, otherwise using the Pearson chi-squared test.

Finally the package *experDesign*, on [CRAN](https://cran.r-project.org/package=experDesign), provides similar functionality conceptually similar to *anticlust* but does accept categorical variables.

The main function `design`distribute the samples on multiple batches so that each variable is equally homogeneous in each batch. If the experiment is carried out on a specific spatial distribution the `spatial` function takes into consideration the position of the samples to distribute them homogeneously also by position.
